VOI-aware Monte Carlo Sampling in Trees

### 2011-12-05 paper/vago/paper.tex
paper in even more progress
paper in progress

\begin{abstract}
Upper bounds for the VOI are provided for pure exploration in the
Multi-armed Bandit Problem. Sampling policies based on the upper
bounds are suggested. Empirical evaluation of the policies is provided
on random problems as well as on the Go game.
\end{abstract}

@@

The intrinsic VOI of pulling an arm is the expected decrease
in the regret compared to selecting an arm without pulling any arm at
all. The intrinsic VOI can be estimated as the intrinsic 
value of perfect information $\Lambda_i$ about the mean reward of the $i$th arm. Two
cases are possible:
\begin{itemize}
\item the arm ($\alpha$) with the highest sample mean is pulled, and the 
mean of the arm is lower than the sample mean the second-best arm ($\beta$);
\item another arm is pulled, and the mean of the arm is higher
than the current highest sample mean ($\alpha$).
\end{itemize}
$\Lambda_i$ can be bounded from above as the probability that a different arm is selected times the
maximum possible increase in the reward:

@@

\begin{thm} The blinkered estimate $\Lambda_i^b$ of intrinsic value of information of sampling
  the $i$th arm for the remaining budget of $N$ samples is bounded from above as
\begin{equation}
  \Lambda_i^b \le \left\{
  \begin{array}{l l}
    \Pr(\overline X_i'\le\overline X_\beta)\overline X_\beta\frac N {N+n_i} < \Pr(\overline X_i'\le\overline X_\beta)\overline X_\beta\frac N   {n_i} & \mbox{if $i=\alpha$} \\
    \Pr(\overline X_i'\ge\overline X_\alpha)(1-\overline X_\alpha)\frac N {N+n_i} < \Pr(\overline X_i'\ge\overline X_\alpha)(1-\overline X_\alpha)\frac N {n_i}&  \mbox{otherwise}
  \end{array} \right.
\label{eqn:thm-be}
\end{equation}
where $\overline X_i'$ is the sample mean of the $i$th arm after $N$ more
samples.
\end{thm}

 The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be}) can be bounded from above using concentration
inequalities. In particular, Lemma~\ref{lemma:hoeffding-prob-bounds} is
based on the Hoeffding inequality (\ref{eqn:conc-hoeffding}):
\begin{lmm} The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be}) are bounded from above as
\begin{eqnarray}
\Pr(\IE[X_i] \le \overline X_\beta|i=\alpha)& \le & \exp(-2 (\overline X_i - \overline X_\beta)^2 n_i)\nonumber\\
\Pr(\IE[X_i] \ge \overline X_\alpha|i\ne\alpha)& \le & \exp(-2 (\overline X_\alpha - \overline X_i)^2 n_i)\nonumber\\
\Pr(\overline X_i' \le \overline X_\beta|i=\alpha)& \le &  2\exp\left(-\frac {2(\overline X_i - \overline X_\beta)^2  (n_i+N) n_i} {(\sqrt n_i + \sqrt {n_i + N})^2}  \right) \le 2\exp\left(-\frac {(\overline X_i - \overline X_\beta)^2 n_i} 2 \right)\nonumber\\
\Pr(\overline X_i' \ge \overline X_\alpha|i\ne\alpha)& \le & 2\exp\left(-\frac {(\overline X_\alpha - \overline X_i)^2 n_i} 2 \right)
\label{eqn:probound-perf-hoeffding}
\end{eqnarray}
\label{lemma:hoeffding-prob-bounds}
\end{lmm}

@@

\subsection{Selecting best arm}

\subsection{Playing Go against UCT}

@@

@@

The bound can be supposedly be improved by selecting a midpoint
$0 < \gamma < \overline X_\beta$ and computing the bound as the sum of two parts:
\begin{itemize}
\item $\overline X_\beta-\gamma$ multiplied by the probability that
  $\mu_\alpha \le \overline X_\beta$;
\item $\overline X_\beta$ multiplied by the probability that $\mu_\alpha\le
  \gamma$.
\end{itemize}.
\[ V_{\overline X=\overline X_\alpha} \le (\overline X_\beta-\gamma)\exp\left(-2n(\overline X_\alpha-\overline X_\beta)^2\right)+\overline X_\beta\exp\left(-2n(\overline X_\alpha-\gamma)^2\right) \]

@@

The minimum of $V^*$ is achieved when $\frac {dV^*} {d\gamma}=0$, that
is, when $\gamma$ is the root of the following equation:
\[ 4\overline X_\beta n(\overline X_\alpha-\gamma)=\exp\left(-2n\left(\frac {\overline X_\alpha-\overline X_\beta}
    {\overline X_\alpha-\gamma}\right)^2\right) \]
If a root in the interval  $0\le\gamma\le\beta$ exists, then the
number of samples is bounded as
\[n\le\frac 1 {4\overline X_\beta(\overline X_\alpha-\overline X_\beta)}\]
by observing that the right-hand side 
is at most $1$ (a negative power), and the left-hand side is at least
$4\overline X_\beta n(\overline X_\alpha-\overline X_\beta)$.
So, the bound can supposedly be improved for smaller values of
$n$. The improvement is more significant when the current best and
second-best sample means are close.

The derivation for the other case
(sampling an item that can be better than the current best) is
obtained by substitution $1-\overline X, 1-\overline X_\alpha, 1-\gamma$ instead of
$\overline X_\alpha, \overline X_\beta, \gamma$:

\[ V_{\overline X\ne\overline X_\alpha} \le (\gamma-\overline X_\alpha)\exp\left(-2n(\overline X_\alpha-\overline
  X)^2\right)+(1-\overline X_\alpha)\exp\left(-2n(\gamma-\overline X)^2\right) \]

@@

A closed-form solution for $\gamma$ cannot be obtained, but given
$\overline X_\alpha, \overline X_\beta, n$, the value of $\gamma$ can be efficiently
computed. It should be determined empirically whether the improved
estimate has justified influence on the performance of the algorithm.

>>>>

\begin{abstract}
Upper bounds for the VOI are provided for pure exploration in the
Multi-armed Bandit Problem. Sampling policies based on the upper
bounds are suggested. Empirical evaluation of the policies and
comparison to the UCB1 and UCT policies is provided
on random problem instances as well as on the Go game.
\end{abstract}

@@

The intrinsic VOI $\Lambda_i$ of pulling an arm is the expected decrease
in the regret compared to selecting an arm without pulling any arm at
all. The \textit{myopic} VOI estimate is of limited applicability to
Monte Carlo sampling, since the effect of a single sample is small,
and the myopic VOI estimate will often be non-positive, resulting in premature
termination of the search. However, $\Lambda_i$ can be estimated as the intrinsic 
value of perfect information $\Lambda_i^p$ about the mean reward of the $i$th arm. Two
cases are possible:
\begin{itemize}
\item the arm ($\alpha$) with the highest sample mean is pulled, and the 
mean of the arm is lower than the sample mean the second-best arm ($\beta$);
\item another arm is pulled, and the mean of the arm is higher
than the current highest sample mean ($\alpha$).
\end{itemize}
$\Lambda_i^p$ can be bounded from above as the probability that a
different arm is selected, multiplied by the
maximum possible increase in the reward:

@@

The search time is finite, and in a simple case the \textit{search
  budget} specified as the maximum number of samples.
An estimate based on the perfect intrinsic VOI does not take in
consideration the remaining number of samples. Given two arms
with the same intrinsic perfect VOI, the VOI
estimate of the arm pulled fewer times so far should be higher.

\begin{dfn} The \textbf{blinkered estimate} of intrinsic VOI information of the
  $i$th arm is the intrinsic VOI of pulling the $i$th arm for the
  remaining budget.
\end{dfn}

\begin{thm} Denote the current number of samples of the $i$th arm as
  $n_i$. The blinkered estimate $\Lambda_i^b$ of intrinsic value of
  information of pulling the $i$th arm for the remaining budget of $N$
  samples is bounded from above as
\begin{equation}
  \Lambda_i^b \le \left\{
  \begin{array}{l l}
    \Pr(\overline X_i'\le\overline X_\beta)\overline X_\beta\frac N {N+n_i} & \mbox{if $i=\alpha$} \\
    \Pr(\overline X_i'\ge\overline X_\alpha)(1-\overline  X_\alpha)\frac N {N+n_i} &  \mbox{otherwise}
  \end{array} \right.
\label{eqn:thm-be}
\end{equation}
where $\overline X_i'$ is the sample mean of the $i$th arm after $n_i+N$ 
samples.
\end{thm}

 The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be}) can be bounded from above using concentration
inequalities. In particular, Lemma~\ref{lemma:hoeffding-prob-bounds} is
based on the Hoeffding inequality (\ref{eqn:conc-hoeffding}):
\begin{lmm} The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be}) are bounded from above as
\begin{eqnarray}
\Pr(\IE[X_i] \le \overline X_\beta|i=\alpha)& \le & \exp(-2 (\overline X_i - \overline X_\beta)^2 n_i)\nonumber\\
\Pr(\IE[X_i] \ge \overline X_\alpha|i\ne\alpha)& \le & \exp(-2 (\overline X_\alpha - \overline X_i)^2 n_i)\nonumber\\
\Pr(\overline X_i' \le \overline X_\beta|i=\alpha)& \le &  2\exp\left(-\frac {2(\overline X_i - \overline X_\beta)^2  (n_i+N) n_i} {(\sqrt n_i + \sqrt {n_i + N})^2}  \right) < 2\exp\left(-\frac {(\overline X_i - \overline X_\beta)^2 n_i} 2 \right)\nonumber\\
\Pr(\overline X_i' \ge \overline X_\alpha|i\ne\alpha)& \le & 2\exp\left(-\frac {(\overline X_\alpha - \overline X_i)^2 (n_i+N) n_i} {(\sqrt n_i + \sqrt {n_i + N})^2} \right) <  2\exp\left(-\frac {(\overline X_\alpha - \overline X_i)^2 n_i} 2 \right)
\label{eqn:probound-perf-hoeffding}
\end{eqnarray}
\label{lemma:hoeffding-prob-bounds}
\end{lmm}

@@

@@

\subsection{Selecting The Best Arm}

\subsection{Playing Go Against UCT}

@@

The bound can be supposedly be improved by selecting a midpoint
$0 < y < \overline X_\beta$ and computing the bound as the sum of two parts:
\begin{itemize}
\item $\overline X_\beta-y$ multiplied by the probability that
  $\mu_\alpha \le \overline X_\beta$;
\item $\overline X_\beta$ multiplied by the probability that $\mu_\alpha\le
  y$.
\end{itemize}.
\[ V_{\overline X=\overline X_\alpha} \le (\overline X_\beta-y)\exp\left(-2n(\overline X_\alpha-\overline X_\beta)^2\right)+\overline X_\beta\exp\left(-2n(\overline X_\alpha-y)^2\right) \]

@@

The minimum of $V^*$ is achieved when $\frac {dV^*} {dy}=0$, that
is, when $y$ is the root of the following equation:
\[ 4\overline X_\beta n(\overline X_\alpha-y)=\exp\left(-2n\left(\frac {\overline X_\alpha-\overline X_\beta}
    {\overline X_\alpha-y}\right)^2\right) \]
If a root in the interval  $0\le y\le\beta$ exists, then the
number of samples is bounded as
\[n\le\frac 1 {4\overline X_\beta(\overline X_\alpha-\overline X_\beta)}\]
by observing that the right-hand side 
is at most $1$ (a negative power), and the left-hand side is at least
$4\overline X_\beta n(\overline X_\alpha-\overline X_\beta)$.
So, the bound can supposedly be improved for smaller values of
$n$. The improvement is more significant when the current best and
second-best sample means are close.

The derivation for the other case
(sampling an item that can be better than the current best) is
obtained by substitution $1-\overline X, 1-\overline X_\alpha, 1-y$ instead of
$\overline X_\alpha, \overline X_\beta, y$:

\[ V_{\overline X\ne\overline X_\alpha} \le (y-\overline X_\alpha)\exp\left(-2n(\overline X_\alpha-\overline
  X)^2\right)+(1-\overline X_\alpha)\exp\left(-2n(y-\overline X)^2\right) \]

@@

A closed-form solution for $y$ cannot be obtained, but given
$\overline X_\alpha, \overline X_\beta, n$, the value of $y$ can be efficiently
computed. It should be determined empirically whether the improved
estimate has justified influence on the performance of the algorithm.
